\fancyhf{}
\fancyhead[C]{Chapter 8. Appendices}% <- added
\fancyfoot[R]{\thepage\ifodd\value{page}\else\hfill\fi}
%\fancyhead[L]{\ifodd\value{page}\relax\else\hfill\fi Ch \thechapter}
%\renewcommand\headrulewidth{0pt}% default ist .4pt
\renewcommand{\plainheadrulewidth}{.4pt}% default is 0pt
\newpage


\section{Acknowledgements}
\begin{spacing}{1.5}
\tab It is difficult to describe how I feel about everyone who has been with me in this journey towards getting my doctorate degree. However, I will do my best to put my feelings into words!

First and foremost, thank you \textbf{Carolina} for giving me the opportunity to perform research in your lab! Although you might be notoriously famous for your ``RBF" ;), I felt that it did not represent how you are as a promotor and a person. Your compassionate guidance and commitment to the improvement of your students have shaped me into the scientist that I am today. Your eye for detail, scientific knowledge, and critical thinking showed me that I have had a professional and experienced promotor on my side. I am fortunate to have had you as my mentor, or rather, a science mom. :)

\textbf{Roseriet}, as my co-promotor, you have been a gigantic help throughout this entire journey. Details on projects were easy for you to remember, which made me feel at ease knowing that you were extremely involved in all phases of my PhD track. Your meticulous mind and pragmatic thinking were a great combination of skills that helped me make important decisions. Thank you! Both you and Carolina are hard workers, and you both have also taught me the value of work-life balance. I think I have grown the most in my life during the years that I was under both your supervision, on a professional as well as personal level.

\textbf{Alejandro}, as my co-promotor, you've been supportive of me since the start of my PhD track. Although our contact was limited, your support came at the right times when situations seemed challenging, or even gloomy. I am grateful for the opportunities you facilitated for me to engage with colleagues both within and beyond the Radboud community. The experience in Rome was unforgettable.

Major gratitude goes to \textbf{every single participant of the BINGO and BIBO study}! The completion of this dissertation would have been impossible without your commitment, excitement, and dedication. BINGO was such a wonderful first experience with human data collection! Every single toddler was a pleasure to work with. During the COVID pandemic, my life took a somber turn as I missed the human interactions. I therefore felt truly fortunate to have the opportunity to visit all BIBO participants, with each one of you adding an extra touch of sunshine to my life.

Thanks to the other co-authors of my papers, \textbf{Henk Schols}, \textbf{Fangjie Gu}, \textbf{Clara Belzer}, and \textbf{Hauke Smidt}, whose expertise and input were extremely valuable in guiding our research and in understanding what our data was showing us and how it could be interpreted. Special thanks go to \textbf{Jeanne de Vries} and \textbf{Corine Perenboom} for their advice and help regarding the assessment of dietary intake.

My PhD journey started at the Developmental Psychology department at the Radboud University in the first 2 years. I have had warm interactions with many who are already continuing their professional journey. \textbf{Christine}, your initiation of the BINGO study set the stage for an incredible journey. Thank you so much! \textbf{Bonnie}, your quirkiness was refreshing. You showed me that being yourself is the strongest asset one can express. \textbf{Pamela}, you are a kind-hearted colleague whose calmness and knowledge were inspirational. \textbf{Kelly}, as you always provided a listening ear, your kindness never went unnoticed. It was always good to see you. \textbf{Maartje}, of the few interactions that we've had, you struck me as a kind and gentle person. Thanks for handing BIBO over to me :) \textbf{Anouk}, thank you for your advice and knowledge about executive functions, and the shared stories. 
I would also like to thank \textbf{Yvonne van den Berg}, \textbf{Toon Cillessen}, \textbf{Bill Burk}, \textbf{Tessa Lansu}, \textbf{Remco}, \textbf{Sterre}, \textbf{Nicole}, \textbf{Jesse}, \textbf{Ilja}, \textbf{Nina}, \textbf{Lisan}, and \textbf{Geert} for making me feel welcome at the department at the Radboud University (and thanks to the latter 3 for being my brief tennis partners!). I also want to thank \textbf{Brianna}, \textbf{Daria}, and \textbf{Esther} for including me in their activities during the first year of my PhD (I still remember the tasty pumpkin bread!). Special thanks to \textbf{Julian} who helped me create a behavioural task in Psychopy for the BIBO study. \textbf{Peiying}, I always felt so comfortable with you talking about anything, and I had a lot of fun playing tennis with you! Hope you're doing well! \textbf{Kris}, thank you for your major support during this PhD journey!

My PhD journey continued at the Radboudumc. As our lab expanded, I had the pleasure of encountering many kind and intelligent people. \textbf{Irene}, my Paranymph, you were the first person that I felt connected to in our new-forming lab. Your presence, your humour, and your accent (:P) made me smile! I am delighted that we could share our hardships and learnings in the past years. :) \textbf{Hellen}, my Paranymph, radiates a contagious brightness. I loved our conversations about science, sustainability, and more personal topics. I admire you and have learned invaluable lessons from our chats. :) \textbf{Yang}, my precious colleague and co-author, collaborating on our microbiota paper was a pleasure (despite the challenges we encountered!). Your admirable strength and kindness shine through, and I cherish the conversations we shared. Wishing you all the best in China! \textbf{Kathi}, it was so nice to have you as a colleague because of your bright and kind personality. I was happy to hear that you are pursuing something you truly enjoy. \textbf{Nicole}, you are the gentlest and kindest person I know. I very much enjoyed our random dinners together, and your BIBO knowledge tests, haha! I wish you all the best for your PhD! \textbf{Henrik}, you are intelligent, with a sense of humour and a down-to-earth nature. I am very thankful that I could always ask you microbiota-related questions and receive the most well-constructed answers. \textbf{Stefania}, you are such an intelligent and passionate person, and that makes you shine! It was so nice collaborating with you on our caregiving paper :) \textbf{Nina}, I can always count on you being at the office ;) I enjoyed our chats (and rants :P) about our PhD lives during lunch. \textbf{Evie}, I admire your strong, righteous, and gentle personality. It was always a pleasure to chat with you :) \textbf{Ana}, though we were situated in different buildings, it was always good to see you in our lab meetings and at other meet ups :) \textbf{Marjorie} and \textbf{Dafna}, we have not seen each other that much, but it was always fun and pleasant to chat with you!

Other colleagues, students, and research assistants have played a major role in my PhD trajectory! Special thanks to \textbf{Jana}, \textbf{Rosa}, and \textbf{Ilse} who spent a large amount of time helping me with the home visits, and data coding for BINGO and BIBO! You made my data collection time so much smoother, and much more fun! Thank you \textbf{Anna}, for your help setting up the BIBO 12.5 years round. It was a reassurance knowing that we had a brain expert by our side :). \textbf{Timothy}, thanks for your help with programming BIBO's curiosity task! Thank you \textbf{Els}, \textbf{Lisanne}, \textbf{Eefje}, \textbf{Annefleur}, \textbf{Ismay}, \textbf{Noortje}, \textbf{Hilde}, \textbf{Jasmijn}, \textbf{Danique}, and \textbf{Lonneke} for your help with picking up samples, processing samples, preparing materials and/or entering and coding data for BINGO and BIBO, and the great conversations! Thanks to \textbf{Sofia} for helping with the creation of the BIBOok ;D It was such a comforting feeling to have you all by my side!

Other people walking around at the Donders were a major part of how I experienced my work life. \textbf{Vab}, \textbf{Casper} (/former Nijmegen tennis partner), \textbf{Carolina} (/moving partner), \textbf{Alejandro}, \textbf{Mariana}, \textbf{Sevgi}, \textbf{Lea}, \textbf{James}, \textbf{Emma} (/coffee break partner), \textbf{Michael}, \textbf{Claire}, \textbf{Angelos}, and \textbf{Shery} you have all been such great mood-makers during my time at the CNS department. Thank you all for the cosy lunches, shared food, and good/random/silly conversations. :)
\textbf{Jo} and \textbf{Kübra}, two of my former `Best Office Ever' mates, together with Irene, were such pleasant office mates to be around! Thanks for tolerating my phone calls to participants ;) I really enjoyed our dinners together. Let's do that again sometime!
Of course, I want to thank \textbf{Erna}, \textbf{Inge}, \textbf{Annemarie}, and \textbf{Ellen} at the secretary! You have always been there to help me out. And thanks to \textbf{Nino} for your help with my administrative things (e.g., reminding me to fill in my hours hehe). 
Thanks to \textbf{Andrea }and \textbf{Max} from Betty and Mora for the tasty fresh Italian lunches I could always count on!

I would like to thank everyone outside the Radboud who has supported me during my PhD journey. To \textbf{Nard}, the first to believe in this PhD adventure, thank you for your unwavering belief and support! \textbf{Olga}, our conversations have helped me get through my darker days, thank you! \textbf{Dewi}, our annual life updates feel like a comforting tradition. I'm grateful to have you capturing my major life event. \textbf{Marjolein}, I loved our catch-ups every so often. Your empathy always made me feel comfortable to talk to you about everything. \textbf{Ivy Mok}, I enjoyed our chats and rants about life! I hope you are doing well.

I would like to thank \textbf{my colleagues at Danone Nutricia Research}, as the end my PhD journey partly overlapped with my start at this company within this amazing team. From the first moment, I felt very welcomed by everyone. All of your interest in my PhD research has made me feel recognised and extra proud. So thank you all, and especially \textbf{Raka}, \textbf{Nana}, \textbf{Elly}, \textbf{Loret}, and \textbf{Evelien}. 

My most heartfelt gratitude goes to the closest people in my life, people who know best how this entire PhD journey impacted me. \textbf{Aaron}, we met quite close to the start of my PhD and our conversations have been a constant source of comfort. \textbf{Ivy}, from elementary school age to adulthood, our occasional contact is a cherished comfort I can always count on. \textbf{Liselotte}, since the age of 12, you have always been by my side. I think we both have grown a lot. Thanks for hearing out about my PhD (and other) struggles and joys these past 6 years. \textbf{Maga}, throughout the challenges during this PhD, you have been my rock. Thank you for always being there. Of course, I want to thank \textbf{my parents} for being patient with me, and hearing out all my experiences during the past 6 years (and my entire life!). I'm excited to bring the first Doctor's title into the family. \textbf{Nishant}, you've been there with me at the end of my PhD journey, and have helped me tremendously in ways I cannot describe. Thank you for your unwavering support and love. I love you all!

Thank you all! Dank jullie wel! 謝謝大家!
\end{spacing}