\fancyhf{}
\fancyhead[C]{Chapter 8. Appendices}% <- added
\fancyfoot[R]{\thepage\ifodd\value{page}\else\hfill\fi}
%\fancyhead[L]{\ifodd\value{page}\relax\else\hfill\fi Ch \thechapter}
%\renewcommand\headrulewidth{0pt}% default ist .4pt
\renewcommand{\plainheadrulewidth}{.4pt}% default is 0pt
\newpage
\begin{spacing}{1.5}
\section{Research Data Management Statement}
\subsection{Ethics}
\tab The BINGO and BIBO study were conducted in accordance with the 1964 Declaration of Helsinki and its later amendments. No formal objections were made against the study protocol of the BINGO study (ECSW2014-1003-189, and ECSW-2018-034) and the BIBO study (SW2017-1303-497, SW2017-1303-498, and ECSW-2018-067) by the ethics committee of the Social Science faculty of the Radboud University, Nijmegen, the Netherlands. A written informed consent was obtained from all participants regarding the use and storage of their data. Participants were informed that they may decide to discontinue the study at any moment, without giving a reason. This research was supported by the European Union's Horizon 2020 Eat2beNice grant (728018 to C. de Weerth and A. Arias Vásquez), a Jacobs Foundation Advanced Research Fellowship (to C. de Weerth), and a Netherlands Organization for Scientific Research VENI grant (016.195.197 to R. Beijers), VIDI grant (575-25-009 to C. de Weerth) and VICI grant (016.Vici.185.038 to C. de Weerth).

\subsection{FAIR principles}
\subsubsection{1. Findable}
\tab All data from the BINGO and BIBO study, including the raw, cleaned, and master data are stored at the secure network drive of the Donders Institute for Brain, Cognition, and Behavior. The pseudonymization key is included in a subfolder of the BINGO and BIBO folder. Physical data, indicating paper data, are stored in the locked archive of the department of Cognitive Neuroscience at the Donders Institute for Brain, Cognition, and Behavior. 

\subsubsection{2. Accessible} 
\tab The data and research documentation are only accessible to researchers involved in the BINGO and/or BIBO project, and to the Developmental Psychobiology (DPB) lab manager. The data are not freely available since we did not ask participants for consent to store their data in a public online depository. Researchers may ask for permission for the re-use of the data with a methodologically sound proposal. The proposal to access the BINGO and/or BIBO should be directed to carolina.deweerth@radboudumc.nl. If approved, data requestors need to sign a data transfer agreement or research collaboration agreement, depending on the level of collaboration. The anonymous processed data will be shared with the collaborator, and researchers are asked to analyze the data and/or publish the results within two years. 

\subsubsection{3. Interoperable}
\tab Documentation on the set-up of the BINGO and BIBO study can be found in the BINGO and BIBO folder and published papers. Results are reproducible and interpretable as the data, documentation files, and R scripts for the papers described in the current thesis are stored at the secure network drive of the Donders Institute for Brain, Cognition, and Behavior. 

\subsubsection{4. Reusable}
\tab All data will be stored for at least 15 years from moment of data collection and can be reused within this period, as stated in the informed consent forms.

\subsection{Privacy}
\tab The privacy of the BINGO and BIBO study participants was guaranteed by pseudo-nymization. The pseudonymization key that linked the code to personal data was stored on the secure network drive of the Donders Institute for Brain, Cognition, and Behavior in separate folders for the BINGO and BIBO studies. This key is only accessible to the DPB lab manager, and researchers that are involved in the project depending on their role. To contact participants for new measurement waves, the pseudonymization key was not destroyed since the BINGO study is halted, and may continue, and the BIBO study is currently ongoing. If data is shared, no personal data (e.g., addresses, videos with participant faces) will be shared.
\end{spacing}

